Point-process based representation learning for Electronic Health Records

### 2023-02-09 Manuscript_v1.tex
update

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\markboth{\journalname, VOL. XX, NO. XX, XXXX 2017}
{Author \MakeLowercase{\textit{et al.}}: Point-process based representation learning for Electronic Health Records}
\begin{document}
\title{Point-process based representation learning for Electronic Health Records}
\author{Hojjat Karami, \IEEEmembership{Fellow, IEEE}, Anisoara Ionescu, and David Atienza, \IEEEmembership{Member, IEEE}
\thanks{This paragraph of the first footnote will contain the date on 
which you submitted your paper for review. It will also contain support 
information, including sponsor and financial support acknowledgment. For 
example, ``This work was supported in part by the U.S. Department of 
Commerce under Grant BS123456.'' }
\thanks{The next few paragraphs should contain 
the authors' current affiliations, including current address and e-mail. For 
example, F. A. Author is with the National Institute of Standards and 
Technology, Boulder, CO 80305 USA (e-mail: author@boulder.nist.gov). }
\thanks{S. B. Author, Jr., was with Rice University, Houston, TX 77005 USA. He is 
now with the Department of Physics, Colorado State University, Fort Collins, 
CO 80523 USA (e-mail: author@lamar.colostate.edu).}
\thanks{T. C. Author is with 
the Electrical Engineering Department, University of Colorado, Boulder, CO 
80309 USA, on leave from the National Research Institute for Metals, 
Tsukuba, Japan (e-mail: author@nrim.go.jp).}}

@@

\begin{abstract}
    It is very well accepted that missingness in electronic health records (EHRs) are not at random which is regarded as informative missingness. The clinician’s decision on when to observe lab tests over time can be modeled using point processes. We propose a novel framework based on neural point process to analyze laboratory tests of ICU patients. This framework can take into account additional information for better characterization of conditional intensity function (CIF) as well as better accuracy in prediction of future timestamp and labels.
\end{abstract}

\begin{IEEEkeywords}
Enter keywords or phrases in alphabetical 
order, separated by commas. For a list of suggested keywords, send a blank 
e-mail to keywords@ieee.org or visit the site.
\end{IEEEkeywords}

@@

Writing an effective abstract is an indispensable part of any form of research, since it that can motivate the audience to read follow the rest of the text. In this task, I have analyzed five abstracts from the field of artificial intelligence (AI). The table in section 2 shows a summary of the most important aspects of the analyzed abstracts.

@@

 The key advantage of our proposed model is to combine a  transformer-based event encoder with point process loss with a deep attention module that can handle an irregularly sampled time series for a any downstream prediciton task.

The entire data consists of $N$ samples  $\mathcal{D}=\{\mathcal{O}_{i}=(\mathcal{E}_{i},\mathcal{S}_{i}  )\}_{i=1}^N$ where $\mathcal{E}_{i}$ and $\mathcal{S}_{i}$ represents event and state sequences, respectively.

Each event sequnce consist of $L$ tuples $\mathcal{E}_{i}=\{ (t_j, e_j)
\}_{j=1}^{L_i}$ where $t_j\in \mathbb{R}  $ is the event timestamp and $ e_j $ is the event mark.

Event sequence data consists of N sequences $\{\mathcal{S}_{i}\}_{i=1}^N$, where
 each sequence $\mathcal{S}_{i}$ is a series of $L_i$ events $\{ (t_j, e_j)
 \}_{j=1}^{L_i}$. Here, $e_j$ represents events that could be independent our mutually 
 exclusive occurring at $t_i$.
 \newline
 In addition to the event data, we might have additional information
  $\{\mathcal{D}_{i}\}_{i=1}^N$. Suppose that each state is represented as $\{
  (t_k, v_k, m_k) \}_{k=1}^{P_i}$, consisting of a time value $t_k\in
  \mathbb{R}^{+} $, an observed value  $z_k\in \mathbb{R} $ and a modality
  indicator  $m_k\in \{1, ..., M\} $.

General schematic of TEDAM is depicted in FIG which consists of two separate module: Transformer Event Encoder (TEE) and Deep Attention Module (DAM). TEE encodes timestamp and mark of events using a transformer architecture, while DAM encodes all available information using one-dimensional attention mechanism. Finally, the learned representations are concatenated to be optimized for different loss functions.

\subsection{Event Encoder}
We use a similar transformer architecture [thp] for encoding events with minor modifications. A transformer architecture is capable of .

let . that can be embedded using a trainable embedding matrix.

In the first step, we embed all event marks $E_{emb}=E \times U$ where $E_i \in [L,K] $ is the binary encoding matrix of all event marks (multi-label or multi-class), and  $ W_{emb} \in  \mathbb{R}^{K \times d_{emb} }  $ is the trainable embedding matrix. In the second step, timestamps should be encoded and added to the event embedding, however, we propose to concatenate time encodings that can lead to better characterization of conditional intensity functions. Finally, the input of the transformer encoder will be $x_{emb}=[y(k_j), z(t_j)]$.

Here, we use the standard transformer encoder similar to [vaswani] with masking matrix to prevent the model from looking into the future. we obtain the encoded matrix  $ H=(h_1, ..., h_j, ..., h_L) $ where $h_j$ containes the all available information before occurence of $j-th$ event.

\subsection{bin of Event Encoder}

it is necessary to include temporal information. Similar to the original
positional encoding [vaswani], we use a temporal encoding procedure:

        \end{cases} 
\end{equation}

Here, $d_t \in \mathbb{N} $ is the dimensionality of encoded timestamp and $z
\in \mathbb{R}^{d_t} $ is the embeding vector of timestamp.

Each event mark $e_j$ is projected to a sparse binary vector representation. We
add an embedding layer to achieve a more compact and efficient representation
$emb$. Here, $w$ and $b$ are weights and biases of the embedding layer which can be
learned during network training.

While previous works suggested adding temporal encoding to the embedded events,
we propose to concatenate these two vectors. The effectiveness of this
concatenation is further investigated in the results.

\begin{equation} x_{emb}=[y(k_j), z(t_j)] \end{equation}

Now that the $x_{emb}$ is ready for encoding, it is encoded through a standard
transformer encoder with multiple layers and attention heads.

\begin{equation}
    x_{enc}=TE(x_{emb})    
\end{equation}

\subsection{State Encoder}
Here, we propose a method to incorporate additional information for a
better representation. In healthcare, much data is available from different
modalities such as vital signs and laboratory values.

@@

\begin{align*}
    f^{\prime}(\mathcal{S}_p) = g^{\prime} \left(  \frac{1}{p} \sum_{s_k \in \mathcal{S}_p}  h^{\prime}_{\theta}(s_k) \right)  \\
    K_p=[f^{\prime}(\mathcal{S}_p), s_p]^T W^K   
\end{align*}

Furthermore, we define a query vector $w^q \in \mathbb{R}^d$, which allows the model to summarize different aspects of the dataset via

\begin{equation*}
    e_p=\frac{K_p.w^q}{\sqrt{d}  }
\end{equation*}

Now, the desired attention can be computed as follows:

\begin{equation*}
    a(\mathcal{S}_p,s_k)=\frac{\exp (e_p)}{\sum_{k\leq p} \exp (e_k)  } 
\end{equation*}

@@

\begin{equation*}
    f(\mathcal{S}_p) =     
    g_{\psi}\left(
    \sum_{s_k \in \mathcal{S}_p}  a(\mathcal{S}_p,s_k)h_{\theta}(s_k) 
    \right) 
\end{equation*}

@@

\begin{equation*}
    \begin{cases} 
f^{\prime}(\mathcal{S}_p) = g^{\prime} \left(   
 \frac{1}{p} \sum_{s_k \in \mathcal{S}_p}  h^{\prime}_{\theta}(s_k) 
\right) \\
K_p=[f^{\prime}(\mathcal{S}_p), s_p]^T W^K \\
e_p=\frac{K_p.w^q}{\sqrt{d}  }\\
a(\mathcal{S}_p,s_k)=\frac{\exp (e_p)}{\sum_{k\leq p} \exp (e_k)  } \\
f(\mathcal{S}_p) = \sum_{s_k \in \mathcal{S}_p}  a(\mathcal{S}_p,s_k)h_{\theta}(s_k) \\
    z_p=g_{\psi}\left(         f(\mathcal{S}_p)         \right) \\
    \end{cases} 
\end{equation*}

\subsection{Mark and Time Decoder}

by concatenating encoded events $x_{enc}$ and additional information $f^{\prime}(\mathcal{S}_p)$, we can predict next marks and times as follows:

\begin{equation}
    \hat{e}_{j+1}=MLP([x_j, f^{\prime}(\mathcal{S}_j)])
\end{equation}

\begin{equation}
    \hat{t}_{j+1}=MLP([x_j, f^{\prime}(\mathcal{S}_j)])
\end{equation}

\begin{equation}
    \mathcal{L}_{mark} =
\end{equation}
\begin{equation}
    \mathcal{L}_{time} = \sum_{j=1}^{L-1} \left(  (\hat{t}_{j+1}-t_j)-(t_{j+1}-t_j)  \right)^2
\end{equation}

@@

In neural point process literature, many approaches have been propose to decode either conditional or cumulative intensity function.

\begin{equation}
    \lambda_k(t|\mathcal{H}_t ) = f_k\left(\alpha_k \frac{t-t_j}{t_j}+\mathbf{w}_k^T \mathbf{x}_{enc}(t_j)+\mathbf{y}_k^T \mathbf{s}_{enc}(t_j) +b_k   \right)
\end{equation}

@@

\subsection*{Datasets}

@@

% To show the effectiveness of the proposed method, we further consider the following scenarios:
% \begin{itemize}
%     \item Enc(Event)+Dec(NextEvent): Here
%     \item Enc(Event)+Dec(NextEvent+CIF): Here
%     \item Enc(Event+state)+Dec(NextEvent): Here
%     \item Enc(Event+state)+Dec(NextEvent+CIF): Here
% \end{itemize}

\textbf{MIMIC-IV (\emph{M4})}.
We selected Medical Information Mart for Intensive Care (MIMIC) IV [1], which is a real-world clinical database comprising health data relating to over 40,000 patients admitted to ICU at the Beth Israel Deaconess Medical Center.

\textbf{Synthea(\emph{Syn})}.
We used the Synthea simulator (Walonoski et al., 2018) which generates patient-level EHRs using human expert curated Markov processes. Here, we reused the already processed version of this data by Edgauard.

\textbf{Stackoverflow (\emph{SO})}.
StackOverflow is a question-answering website. The website rewards users with badges to promote engagement in the community, and the same badge can be rewarded multiple times to the same user. We collect data in a two-year period, and we treat each user’s reward history as a sequence. Each event in the sequence signifies receipt of a particular medal.

\subsection*{Scenarios}
To investigate the effectiveness of the proposed method, we consider three input scenarios (TE, DA, TE+DA) as well as three loss functions (next event, CIF, next event+CIF) which would result in nine scenarios.

The first series of experiments are conducted to investigate the advantage of encoding additional information for paramterization of intensity functions. We consider seven scenarios:
Here, the baseline models

To show the effectiveness of time concatenation we report.

\subsection*{Baselines}
we use NEURALTPP that is already developed pipeline by as they already considered a lot of comibinations.

@@

\section{Results and Discussion}
\label{sec:Results and Discussion}

@@

% Table generated by Excel2LaTeX from sheet 'time concat'
\begin{table}[h!]
    \centering
    \caption{Add caption}
      \begin{tabular}{ccccc}
      \toprule
            &       & \multicolumn{2}{c}{shp+mark} &  \\
  \cmidrule{3-4}    Dataset & Metric & concat & sum   & baseline \\
      \midrule
      SO    & NLL   & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
            & AUROC & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
            & time  & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
      \midrule
      Synthea & NLL   & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
            & AUROC & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
            & time  & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
      \midrule
      ReTweet & NLL   & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
            & AUROC & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
            & time  & 0.55 (0.02) & 0.55 (0.02) & 0.55 (0.02) \\
      \bottomrule
      \end{tabular}%
    \label{tab:addlabel}%
  \end{table}%

@@

\section{Introduction}
\label{sec:introduction}
\IEEEPARstart{T}{his} document is a template for \LaTeX. If you are 
reading a paper or PDF version of this document, please download the 
electronic file, trans\_jour.tex, from the IEEE Web site at \underline
{http://www.ieee.org/authortools/trans\_jour.tex} so you can use it to prepare your manuscript. If 
you would prefer to use LaTeX, download IEEE's LaTeX style and sample files 
from the same Web page. You can also explore using the Overleaf editor at 
\underline
{https://www.overleaf.com/blog/278-how-to-use-overleaf-with-}\discretionary{}{}{}\underline
{ieee-collabratec-your-quick-guide-to-getting-started\#.}\discretionary{}{}{}\underline{xsVp6tpPkrKM9}

If your paper is intended for a conference, please contact your conference 
editor concerning acceptable word processor formats for your particular 
conference.

IEEE will do the final formatting of your paper. If your paper is intended 
for a conference, please observe the conference page limits.

\subsection{Abbreviations and Acronyms}
Define abbreviations and acronyms the first time they are used in the text, 
even after they have already been defined in the abstract. Abbreviations 
such as IEEE, SI, ac, and dc do not have to be defined. Abbreviations that 
incorporate periods should not have spaces: write ``C.N.R.S.,'' not ``C. N. 
R. S.'' Do not use abbreviations in the title unless they are unavoidable 
(for example, ``IEEE'' in the title of this article).

\subsection{Other Recommendations}
Use one space after periods and colons. Hyphenate complex modifiers: 
``zero-field-cooled magnetization.'' Avoid dangling participles, such as, 
``Using \eqref{eq}, the potential was calculated.'' [It is not clear who or what 
used \eqref{eq}.] Write instead, ``The potential was calculated by using \eqref{eq},'' or 
``Using \eqref{eq}, we calculated the potential.''

Use a zero before decimal points: ``0.25,'' not ``.25.'' Use 
``cm$^{3}$,'' not ``cc.'' Indicate sample dimensions as ``0.1 cm 
$\times $ 0.2 cm,'' not ``0.1 $\times $ 0.2 cm$^{2}$.'' The 
abbreviation for ``seconds'' is ``s,'' not ``sec.'' Use 
``Wb/m$^{2}$'' or ``webers per square meter,'' not 
``webers/m$^{2}$.'' When expressing a range of values, write ``7 to 
9'' or ``7--9,'' not ``7$\sim $9.''

A parenthetical statement at the end of a sentence is punctuated outside of 
the closing parenthesis (like this). (A parenthetical sentence is punctuated 
within the parentheses.) In American English, periods and commas are within 
quotation marks, like ``this period.'' Other punctuation is ``outside''! 
Avoid contractions; for example, write ``do not'' instead of ``don't.'' The 
serial comma is preferred: ``A, B, and C'' instead of ``A, B and C.''

If you wish, you may write in the first person singular or plural and use 
the active voice (``I observed that $\ldots$'' or ``We observed that $\ldots$'' 
instead of ``It was observed that $\ldots$''). Remember to check spelling. If 
your native language is not English, please get a native English-speaking 
colleague to carefully proofread your paper.

Try not to use too many typefaces in the same article. You're writing
scholarly papers, not ransom notes. Also please remember that MathJax
can't handle really weird typefaces.

\subsection{Equations}
Number equations consecutively with equation numbers in parentheses flush 
with the right margin, as in \eqref{eq}. To make your equations more 
compact, you may use the solidus (~/~), the exp function, or appropriate 
exponents. Use parentheses to avoid ambiguities in denominators. Punctuate 
equations when they are part of a sentence, as in
\begin{equation}E=mc^2.\label{eq}\end{equation}

Be sure that the symbols in your equation have been defined before the 
equation appears or immediately following. Italicize symbols ($T$ might refer 
to temperature, but T is the unit tesla). Refer to ``\eqref{eq},'' not ``Eq. \eqref{eq}'' 
or ``equation \eqref{eq},'' except at the beginning of a sentence: ``Equation \eqref{eq} 
is $\ldots$ .''

\subsection{\LaTeX-Specific Advice}

Please use ``soft'' (e.g., \verb|\eqref{Eq}|) cross references instead
of ``hard'' references (e.g., \verb|(1)|). That will make it possible
to combine sections, add equations, or change the order of figures or
citations without having to go through the file line by line.

Please don't use the \verb|{eqnarray}| equation environment. Use
\verb|{align}| or \verb|{IEEEeqnarray}| instead. The \verb|{eqnarray}|
environment leaves unsightly spaces around relation symbols.

Please note that the \verb|{subequations}| environment in {\LaTeX}
will increment the main equation counter even when there are no
equation numbers displayed. If you forget that, you might write an
article in which the equation numbers skip from (17) to (20), causing
the copy editors to wonder if you've discovered a new method of
counting.

{\BibTeX} does not work by magic. It doesn't get the bibliographic
data from thin air but from .bib files. If you use {\BibTeX} to produce a
bibliography you must send the .bib files.

{\LaTeX} can't read your mind. If you assign the same label to a
subsubsection and a table, you might find that Table I has been cross
referenced as Table IV-B3.

{\LaTeX} does not have precognitive abilities. If you put a
\verb|\label| command before the command that updates the counter it's
supposed to be using, the label will pick up the last counter to be
cross referenced instead. In particular, a \verb|\label| command
should not go before the caption of a figure or a table.

Do not use \verb|\nonumber| inside the \verb|{array}| environment. It
will not stop equation numbers inside \verb|{array}| (there won't be
any anyway) and it might stop a wanted equation number in the
surrounding equation.

If you are submitting your paper to a colorized journal, you can use
the following two lines at the start of the article to ensure its
appearance resembles the final copy:

\smallskip\noindent
\begin{small}
\begin{tabular}{l}
\verb+\+\texttt{documentclass[journal,twoside,web]\{ieeecolor\}}\\
\verb+\+\texttt{usepackage\{\textit{Journal\_Name}\}}
\end{tabular}
\end{small}

\section{Units}
Use either SI (MKS) or CGS as primary units. (SI units are strongly 
encouraged.) English units may be used as secondary units (in parentheses). 
This applies to papers in data storage. For example, write ``15 
Gb/cm$^{2}$ (100 Gb/in$^{2})$.'' An exception is when 
English units are used as identifiers in trade, such as ``3\textonehalf-in 
disk drive.'' Avoid combining SI and CGS units, such as current in amperes 
and magnetic field in oersteds. This often leads to confusion because 
equations do not balance dimensionally. If you must use mixed units, clearly 
state the units for each quantity in an equation.

The SI unit for magnetic field strength $H$ is A/m. However, if you wish to use 
units of T, either refer to magnetic flux density $B$ or magnetic field 
strength symbolized as $\mu _{0}H$. Use the center dot to separate 
compound units, e.g., ``A$\cdot $m$^{2}$.''

\section{Some Common Mistakes}
The word ``data'' is plural, not singular. The subscript for the 
permeability of vacuum $\mu _{0}$ is zero, not a lowercase letter 
``o.'' The term for residual magnetization is ``remanence''; the adjective 
is ``remanent''; do not write ``remnance'' or ``remnant.'' Use the word 
``micrometer'' instead of ``micron.'' A graph within a graph is an 
``inset,'' not an ``insert.'' The word ``alternatively'' is preferred to the 
word ``alternately'' (unless you really mean something that alternates). Use 
the word ``whereas'' instead of ``while'' (unless you are referring to 
simultaneous events). Do not use the word ``essentially'' to mean 
``approximately'' or ``effectively.'' Do not use the word ``issue'' as a 
euphemism for ``problem.'' When compositions are not specified, separate 
chemical symbols by en-dashes; for example, ``NiMn'' indicates the 
intermetallic compound Ni$_{0.5}$Mn$_{0.5}$ whereas 
``Ni--Mn'' indicates an alloy of some composition 
Ni$_{x}$Mn$_{1-x}$.

\begin{figure}[!t]
\centerline{\includegraphics[width=\columnwidth]{fig1.png}}
\caption{Magnetization as a function of applied field.
It is good practice to explain the significance of the figure in the caption.}
\label{fig1}
\end{figure}

Be aware of the different meanings of the homophones ``affect'' (usually a 
verb) and ``effect'' (usually a noun), ``complement'' and ``compliment,'' 
``discreet'' and ``discrete,'' ``principal'' (e.g., ``principal 
investigator'') and ``principle'' (e.g., ``principle of measurement''). Do 
not confuse ``imply'' and ``infer.''

Prefixes such as ``non,'' ``sub,'' ``micro,'' ``multi,'' and ``ultra'' are 
not independent words; they should be joined to the words they modify, 
usually without a hyphen. There is no period after the ``et'' in the Latin 
abbreviation ``\emph{et al.}'' (it is also italicized). The abbreviation ``i.e.,'' means 
``that is,'' and the abbreviation ``e.g.,'' means ``for example'' (these 
abbreviations are not italicized).

A general IEEE styleguide is available at \underline{http://www.ieee.org/authortools}.

\section{Guidelines for Graphics Preparation and Submission}
\label{sec:guidelines}

\subsection{Types of Graphics}
The following list outlines the different types of graphics published in 
IEEE journals. They are categorized based on their construction, and use of 
color/shades of gray:

\subsubsection{Color/Grayscale figures}
{Figures that are meant to appear in color, or shades of black/gray. Such 
figures may include photographs, illustrations, multicolor graphs, and 
flowcharts.}

\subsubsection{Line Art figures}
{Figures that are composed of only black lines and shapes. These figures 
should have no shades or half-tones of gray, only black and white.}

\subsubsection{Author photos}
{Head and shoulders shots of authors that appear at the end of our papers. }

\subsubsection{Tables}
{Data charts which are typically black and white, but sometimes include 
color.}

>>>>

\usepackage{biblatex} %Imports biblatex package
% \addbibresource{sample.bib} %Import the bibliography file

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\markboth{\journalname, VOL. XX, NO. XX, XXXX 2017}
{Author \MakeLowercase{\textit{et al.}}: Point-process based representation learning for Electronic Health Records}
\begin{document}
\title{Point-process based representation learning for Electronic Health Records}
\author{Hojjat Karami, \IEEEmembership{Fellow, IEEE}, Anisoara Ionescu, and David Atienza, \IEEEmembership{Member, IEEE}
\thanks{This paragraph of the first footnote will contain the date on 
which you submitted your paper for review. It will also contain support 
information, including sponsor and financial support acknowledgment. For 
example, ``This work was supported by DigiPredict Grant BS123456.'' }
\thanks{H. Karami is with the EPFL, Lausanne, Switzerland (e-mail: hojjat.karami@epfl.ch). }
\thanks{S. B. Author, Jr., was with Rice University, Houston, TX 77005 USA. He is 
now with the Department of Physics, Colorado State University, Fort Collins, 
CO 80523 USA (e-mail: author@lamar.colostate.edu).}
\thanks{T. C. Author is with 
the Electrical Engineering Department, University of Colorado, Boulder, CO 
80309 USA, on leave from the National Research Institute for Metals, 
Tsukuba, Japan (e-mail: author@nrim.go.jp).}}

@@

\begin{abstract}

    Irregular @hornSetFunctionsTime2020 sampling of time series in electronic health records (EHRs) is challenging for model development. In addition, the pattern of missingness for certain clinical variables are not at random as it is determined by clinicians decision and the state of patient. Point process is a mathematical framework for handling event sequence data which is also consistent with the irregular sampling. We propose \emph{TEEDAM}, a deep neural network that can learn patient representation from irregulary sample time series as well as informative missingness pattern of certain laboratory variables. We performed various experiments to show the effectiveness of event and state encoding for characterization of conditional intensity functions as well as downstream prediction task. Results show that in some cases learning from patterns may improve the performance of prediction task.
\end{abstract}

\begin{IEEEkeywords}
    electronic health records (EHRs), Point Process, irregular sampling, informative missingness
\end{IEEEkeywords}

@@

healthcare systems are one of main area for deploying machine learning models. Thanks to recent advancements in data collection technologies, policies and computing power data are recorded in a hosptial. This data consists of multiple modalities frome tabular and time series to clinical notes and images. We focus on tabular and time series.

talk about the irregular sampling in EHRs and causes.///
In and ideal setup, variables are measured on a regular basis, however, in a hospital due to medical costs, patient associated risks, sampling device.

conseq of irr sampling///

potential adv of irr sampl///

imputation methods///

inherently compatible methods///

pp framework///

Our aims??// In this work, we propose a DNN for joint modeling of NTPP and a DAM. Our first aim is to investigate the effectiveness of state encoding for CIF characterization. Second, we are interested in seeing the utility of event encoding a downstream task

contribs?///

@@

 The key advantage of our proposed model is to combine a  transformer-based event encoder (TEE) with point process loss with a deep attention module (DAM) that can handle an irregularly sampled time series for a any downstream prediciton task. In this situation the data is represented as $\mathcal{D}=\{\mathcal{S}_i,\mathcal{U}_i\}_{i=1}^{N}$. General schematic of TEDAM is depicted in FIG.

\subsection{Event Encoder}
We use a similar transformer architecture [thp] for encoding events with minor modifications for event embedding.

In the first step, we embed all event marks $E_{emb}=E \times W_emb$ where $E_i \in \mathbb{R}^{L \times M}$ is the binary encoding matrix of all event marks (multi-label or multi-class), and  $ W_{emb} \in  \mathbb{R}^{M \times d_{emb} }  $ is the trainable embedding matrix.

In the second step, timestamps should be encoded and added to the event embedding, however, we propose to concatenate time encodings that can lead to better characterization of conditional intensity functions. Finally, the input of the transformer encoder will be $X_{emb}=[E_{emb}, T_{emb}] \in \mathbb{R}^{L \times (d_emb+t_emb)}$.

Here, we use the standard transformer encoder similar to [vaswani] with masking matrix to prevent the model from looking into the future. we obtain the encoded matrix  $ H=(h_1, ..., h_j, ..., h_L) $ where $h_j$ containes the all available information before occurence of $j$-th event.

\subsection{State Encoder}

@@

The goal is to calculate $a(\mathcal{S}_p,s_k ), k \leq p$ that is the relevance of $k$-th observation $s_k$ to the first $p$ observed values $\mathcal{S}_p$.

This is achieved by embedding set elements $\mathcal{S}_p$ using a simpler set function $f^{\prime}$ such as a cumulative mean.

\begin{equation}
    f^{\prime}(\mathcal{U}_p) = g^{\prime} \left(  \frac{1}{|p|} \sum_{u_k \in \mathcal{U}_p}  h^{\prime}(u_k;\theta^{\prime}) ;\rho^{\prime}\right) 
\end{equation}

Then we can compute key values using key matrix $W^k \in \mathbb{R}^{(d_g+d_s) \times d_{prod}}$

\begin{equation}
    K_p=[f^{\prime}(\mathcal{U}_p), u_p]^T W^K   
\end{equation}

using a query vector $w^q \in \mathbb{R}^{d_{prod}}$

\begin{equation*}
    a(\mathcal{U}_p,u_p)=  softmax(\frac{K_p.w^q}{\sqrt{d}  })
\end{equation*}

@@

\begin{equation*}
    z_p = f(\mathcal{U}_p) =     
    g\left(
    \sum_{s_k \in \mathcal{U}_p}  a(\mathcal{U}_p,s_k)h(s_k;\theta);\rho 
    \right) 
\end{equation*}

we regard $z_p \in \mathbb{R}^{d_{\rho}}$ as the representation of state data until arrival of $p$-th datapoint.

Finally, we need to combine event embeddings $ H_{L \times d_e}$ and state embeddings $Z_{P \times d_{\rho}}$, however, the length of each does not match. As a result, we consider the reduced version of state matrix as below:

\begin{equation*}
    Z^{red}_i=Z_p where p=argmax(z \leq i)
\end{equation*}

@@

\begin{equation*}
    \begin{cases}

f^{\prime}(\mathcal{S}_p) = g^{\prime} \left(   
 \frac{1}{p} \sum_{s_k \in \mathcal{S}_p}  h^{\prime}_{\theta}(s_k) 
\right) \\

K_p=[f^{\prime}(\mathcal{S}_p), s_p]^T W^K \\
e_p=\frac{K_p.w^q}{\sqrt{d}  }\\

a(\mathcal{S}_p,s_k)=\frac{\exp (e_p)}{\sum_{k\leq p} \exp (e_k)  } \\

f(\mathcal{S}_p) = \sum_{s_k \in \mathcal{S}_p}  a(\mathcal{S}_p,s_k)h_{\theta}(s_k) \\

z_p=g_{\psi}\left(         f(\mathcal{S}_p)         \right) \\

\end{cases} 
\end{equation*}

        \end{cases} 
\end{equation} 
\begin{equation}
    \lambda_k(t|\mathcal{H}_t ) = f_k\left(\alpha_k \frac{t-t_j}{t_j}+\mathbf{w}_k^T \mathbf{x}_{enc}(t_j)+\mathbf{y}_k^T \mathbf{s}_{enc}(t_j) +b_k   \right)
\end{equation}

@@

In neural point process literature, many approaches have been propose to decode either conditional or cumulative intensity function. We will use a decoder similar to [sahp] as it can model both exciting and inhibiting effects for modeling CIFs.

% \begin{gather*} 
%    \mu_{m,i+1}=gelu(h_{i+1}W_{m,\mu}), 
% \end{gather*}

\begin{gather*} 
    \mu_{m,i+1}=gelu(h_{i+1}W_{m,\mu}  +  z_{i+1}W_{m,\mu}), \\ 
    \eta_{m,i+1}=gelu(h_{i+1}W_{m,\eta}  +  z_{i+1}W_{m,\eta}), \\
    \gamma_{m,i+1}=gelu(h_{i+1}W_{m,\gamma}  +  z_{i+1}W_{m,\gamma}),
    \end{gather*}

Finally, we can express the intensity function as follows:

\begin{multline*}
    \lambda_m(t)=softplus(\mu_{m,i+1}+\\
    (\eta_{m,i+1}- \mu_{m,i+1}) \exp(-\gamma_{m,i+1}(t-t_i))    ),
\end{multline*}

for $t \in (t_i, t_{i+1}]$,where the $softplus$ is used to constrain thei intensity function to be positive.

\subsection{Loss Function}

We define a multi-objective loss function 
$ \mathcal{L}   =   \mathcal{L}_{CIF}   +   \mathcal{L}_{mark}  +   \mathcal{L}_{state}$.

@@

We perform various experiments to show the effectiveness of each component in out model.
\subsection*{Datasets}

To show the utility of time concatenation and marked loss, we consider three datasets:

\textbf{Synthea(\emph{Syn})}.
We used the Synthea simulator (Walonoski et al., 2018) which generates patient-level EHRs using human expert curated Markov processes. Here, we reused the already processed version of this data by [ntpp].

\textbf{ReTweets (\emph{RT})}.
The Retweets dataset contains sequences of tweets, where each sequence contains an origin tweet (i.e., some user initiates a tweet), and some follow-up tweets. We record the time and the user tag of each tweet. Further, users are grouped into three categories based on the number of their followers: “small”, “medium”, and “large”

\textbf{Stackoverflow (\emph{SO})}.
is a question-answering website. The website rewards users with badges to promote engagement in the community, and the same badge can be rewarded multiple times to the same user. We collect data in a two-year period, and we treat each user’s reward history as a sequence. Each event in the sequence signifies receipt of a particular medal.

Furthermore, we consider two EHRs provided by physionet challenge to investigate the advantage of irregular sample and point process modeling in the same time.

\textbf{Physionet 2012 Mortality Prediction Challenge (\emph{P19})}.
The 2012 Physionet challenge dataset (Goldberger et al., 2000),contains 12, 000 ICU stays each of which lasts at least 48 h. For each stay, a set of general descriptors (such as gender or age) are collected at admission time. Depending on the course of the stay and patient status, up to 37 time series variables were measured (e.g. blood pressure, lactate, and respiration rate). While some modalities might be measured in regular time intervals (e.g. hourly or daily), some are only collected when required; moreover, not all variables are available for each stay.

@@

\subsection*{Scenarios}

To show the utility of time concatenation and marked-shp we report the metrics for SO, RT and SYN and compare it with 3 baselines: SAHP, THP, and GRU-CP.

To show the utility of additional information for CIF modeling, we report NLL/events for P12 and P19 in the following conditions: TE, TE+DAM, TE+noise

Finally, we investigate whether point process characterization can be useful in a downstream task or not. As each dataset consists of different hospitals and the pattern of irregular sampling might differ from hosptial to hospital, we assume different settings: single-center, multi-center and external evaluation.

\subsection*{Baselines}

we compare our model against existing models: THP, SAHP, GRU-CP.

@@

\subsection*{Training Details}
To be completed.
\section{Results and Discussion}
\label{sec:Results and Discussion}

@@

% Table generated by Excel2LaTeX from sheet 'time concat'
\begin{table*}[htbp]
    \centering
    \caption{Add caption}
      \begin{tabular}{cccccc}
      \toprule
            &       & \multicolumn{2}{c}{TEEDAM} &       &  \\
  \cmidrule{3-4}    Dataset & Metric & concat & sum   & SAHP  & GRU-CP \\
      \midrule
      \multirow{2}[2]{*}{SO} & LL/\#events &       &       &       &  \\
            & AUROC &       &       &       &  \\
      \midrule
      \multirow{2}[2]{*}{Synthea} & LL/\#events &       &       &       &  \\
            & AUROC &       &       &       & 0.85(.014) \\
      \midrule
      \multirow{2}[2]{*}{ReTweet} & LL/\#events & 1.83 (0.09) & 1.5 (0.18) &       &  \\
            & AUROC & 0.68 (0.03) & 0.67 (0.05) &       & 0.611(0.001) \\
      \bottomrule
      \end{tabular}%
    \label{tab:addlabel}%
  \end{table*}%

@@
